Nonlinear Dendritic Coincidence Detection for Supervised Learning

### 2021-05-25 notes/frontiers_coincidence_learning/frontiers.tex
.

%---------------------------------------%
\subsection{Alignment between Basal and Apical Inputs}
\label{sect:alignment}
%---------------------------------------%

@@

@@

For a comparison, in 
Fig.~\ref{fig_corr_dimension_scaling}
data for both the compartment model
and for a point neuron are presented,
as defined respectively by
\eqref{eq_comp_model}
and \eqref{eq_point_neuron}.
One can observe a decorrelation transition for 
both models. However, the compartment model supports a 
significantly stronger distraction in terms of the scaling 
factor $s$ as compared to the point model. This is a first confirmation 
of our hypothesis that nonlinear interactions between basal and 
apical input could improve learning guided by top-down signals.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\centering
\includegraphics[width=1.0\columnwidth]{corr_dimension_scaling_high_input_dim}
\caption{{\bf Alignment between Basal and Apical Input.} Color
encoded the Pearson correlation $\rho[I_p,I_d]$ 
between the proximal and distal input currents,
$I_p$ and $I_d$. Data for a range of the orthogonal 
distraction directions, $N_{\rm dist}\in[0,N-1]$, 
and the scaling factor $s$, as defined in
Fig.~\ref{fig_illustration_classification}.
The overall number of basal inputs is $N=100$.
The quantity $\Sigma_{\rho}$ depicted in the bar
plot is the sum over all correlation values for the different tested values
of $s$ under a given $N_{\rm dist}$. Blue bar represents the compartment model,
orange the point model.}
\label{fig_corr_dimension_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Next, we investigated if the observed differences would also improve
the performance in an actual supervised learning task.
For this purpose, we constructed presynaptic basal 
input as illustrated in Fig.~\ref{fig_illustration_classification}.
Written in vector form, each sample from the basal 
input is generated from the following expression:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\mathbf{x}_p(t) = \mathbf{b} + \mathbf{a}\left(c(t) + \sigma_a \zeta_a(t) \right) 
+ s \cdot \sum_{i=1}^{N_{\rm dist}} \zeta_{dist,i}(t) \mathbf{v}_{dist,i} \; .
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, $\mathbf{b}$ is a random vector drawn uniformly from
$(0,1)^N$, $\mathbf{a}$ is random unit vector as introduced in 
Section~\ref{sect:alignment}, $c(t)$ is a binary variable drawn 
from $\{-0.5,0.5\}$ with equal probability and $\zeta_a(t)$ and the
$\zeta_{dist,i}(t)$ are independent random Gaussian variables with 
zero mean and unit variance. 
Hence, $\sigma_a$ simply denotes the standard deviation of each Gaussian
cluster along the direction of the normal vector $\mathbf{a}$ and was
set to $\sigma_a = 0.25$. 
Finally, the set of $\mathbf{v}_{dist,i}$ forms a randomly generated
orthogonal basis of $N_{\rm dist}$ unit vectors which---just as in 
Section~\ref{sect:alignment}---are also orthogonal to $\mathbf{a}$.
The free parameter $s$ parameterized
the standard deviation along this subspace orthogonal to $\mathbf{a}$.
As indicated by the time dependence, the Gaussian and binary random
variables are drawn for each time step. The vectors
$\mathbf{b}$, $\mathbf{a}$, and $\mathbf{v}_{dist,i}$ are generated
once before the beginning of a simulation run.

@@

The resulting accuracy as a function of $N_{\rm dist}$ and $s$
is shown in Fig.~\ref{fig:classification_accuracy}.
Albeit differences are small, the compartment model does
show a better overall accuracy for the tested parameter range.
It should be noted, though, that the advantage of the compartment
model becomes even less prominent when looking at the actual
correlation between proximal and distal input as a
measure of successful learning (as done in the previous section).
Fig.~\ref{fig:classification_correlation} in the supplementary 
section shows that the compartment model allows for a marginally
larger scaling among the distraction dimensions.
\\
%---------------------------------------%
\subsection{Non-Hebbian Learning Rules}
\label{sect:non-hebbian}
%---------------------------------------%

Instead of Hebbian learning, we also considered a BCM-like
learning rule for the basal weights \citep{Bienenstock1982,Intrator1992}.
The form of the BCM-rule we consider here reads
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\Delta w_i \propto y\left(y - \theta_M\right) x_i - 
\epsilon w_i \; , \label{eq:bcm_rule}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\theta_M$ is a threshold defining a transition from LTP to LTD and
$\epsilon$ is an optional decay term on the weights.
In the variant introduced by \citet{Law1994}, the sliding threshold is simply
the temporal average of the squared neural activity, 
$\theta_M = \langle y^2 \rangle$. In practice, this would be calculated
as a running average, thereby preventing the weights from growing 
indefinitely.

>>>>

%---------------------------------------%
\subsection{Unsupervised Alignment between Basal and Apical Inputs}
\label{sect:alignment}
%---------------------------------------%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\centering
\includegraphics[width=1.0\columnwidth]{corr_dimension_scaling_high_input_dim}
\caption{{\bf Unsupervised Alignment between 
Basal and Apical Input.} Color
encoded the Pearson correlation $\rho[I_p,I_d]$ 
between the proximal and distal input currents,
$I_p$ and $I_d$. Data for a range of the orthogonal 
distraction directions, $N_{\rm dist}\in[0,N-1]$, 
and the scaling factor $s$, as defined in
Fig.~\ref{fig_illustration_classification}.
The overall number of basal inputs is $N=100$.
The quantity $\Sigma_{\rho}$ depicted in the bar
plot is the sum, for a given $N_{\rm dist}$,
over the correlation values $s=0,\,0.5,\,1.0\,..$ 
used for the simulation run. Blue bars represents 
the compartment model, orange the point model.}
%\Delta s = 10/(N_s-1) von 0 bis 10.
% > D.h.  s\in[0,10],   N_s = 20 ?
\label{fig_corr_dimension_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

For a comparison, in 
Fig.~\ref{fig_corr_dimension_scaling}
data for both the compartment model
and for a point neuron are presented,
as defined respectively by \eqref{eq_comp_model}
and \eqref{eq_point_neuron}. A decorrelation 
transition as a function of the distraction 
scaling paramerer $s$ is observed for both 
models. However, the compartment model is able
to handle a significantly stronger distraction 
as compared to the point model. These findings
support the hypothesis examined here, namely
that nonlinear interactions between basal and 
apical input improve learning guided by top-down signals.

@@

Next, we investigated if the observed differences would also improve
the performance in an actual supervised learning task.
For this purpose, we constructed presynaptic basal 
input $x_p(t)$ as illustrated in 
Fig.~\ref{fig_illustration_classification}.
Written in vector form, each sample from the basal 
input is generated from the following expression:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\mathbf{x}_p(t) = \mathbf{b} + \mathbf{a}\left(c(t) + \sigma_a \zeta_a(t) \right) 
+ s \cdot \sum_{i=1}^{N_{\rm dist}} \zeta_{dist,i}(t) \mathbf{v}_{dist,i} \; .
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, $\mathbf{b}$ is a random vector drawn uniformly from
$(0,1)^N$, $\mathbf{a}$ is random unit vector as introduced in 
Section~\ref{sect:alignment}, $c(t)$ is a binary variable drawn 
from $\{-0.5,0.5\}$ with equal probability and $\zeta_a(t)$ and the
$\zeta_{dist,i}(t)$ are independent random Gaussian variables with 
zero mean and unit variance. 
Hence, $\sigma_a$ simply denotes the standard deviation of each Gaussian
cluster along the direction of the normal vector $\mathbf{a}$ and was
set to $\sigma_a = 0.25$. 
Finally, the set of $\mathbf{v}_{dist,i}$ forms a randomly generated
orthogonal basis of $N_{\rm dist}$ unit vectors which---just as in 
Section~\ref{sect:alignment}---are also orthogonal to $\mathbf{a}$.
The free parameter $s$ parameterized
the standard deviation along this subspace orthogonal to $\mathbf{a}$.
As indicated by the time dependence, the Gaussian and binary random
variables are drawn for each time step. The vectors
$\mathbf{b}$, $\mathbf{a}$, and $\mathbf{v}_{dist,i}$ are generated
once before the beginning of a simulation run.

@@

The resulting accuracy as a function of $N_{\rm dist}$ and $s$
is shown in Fig.~\ref{fig:classification_accuracy}.
Albeit differences are small, the compartment model does
show a better overall accuracy for the tested parameter range.
It should be noted, though, that the advantage of the compartment
model becomes even less prominent when looking at the actual
correlation between proximal and distal input as a
measure of successful learning (as done in the previous section).
Fig.~\ref{fig:classification_correlation} in the supplementary 
section shows that the compartment model allows for a marginally
larger scaling among the distraction dimensions.
\\
%---------------------------------------%
\subsection{Generalized Hebbian Learning Rules}
\label{sect:non-hebbian}
%---------------------------------------%

Instead of basic Hebbian learning, we also considered 
a BCM-like learning rule for the basal weights 
\citep{Bienenstock1982,Intrator1992}.
The form of the BCM-rule used here reads
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\Delta w_i \propto y\left(y - \theta_M\right) x_i - 
\epsilon w_i \; , \label{eq:bcm_rule}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\theta_M$ is a threshold defining a transition from LTP to LTD and
$\epsilon$ is an optional decay term on the weights.
In the variant introduced by \citet{Law1994}, the sliding threshold is simply
the temporal average of the squared neural activity, 
$\theta_M = \langle y^2 \rangle$. In practice, this would be calculated
as a running average, thereby preventing the weights from growing 
indefinitely.
